---
title: "Tidy data loading, labelling, summarizing"
knit: (function(input_file, encoding) {
    out_dir <- 'docs';
    rmarkdown::render(input_file,
      encoding=encoding,
      output_file=file.path(dirname(input_file), out_dir, 'index.html'))})
author: "Margaret Janiczek"
date: "2022-12-02"
output: 
  html_document:
    toc: true
    number_sections: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Data is downloaded from: https://thisisstatistics.org/falldatachallenge/. 

Details on the dataset and prompt [here](https://docs.google.com/document/d/1qXnbS4fjuQ9hbRxGaSmix3CGb3FUi6U5/edit)

[Full 2019 codebook describing data](https://nces.ed.gov/nhes/data/2019/pfi/cbook_pfi_pu.pdf)

[Curated dataset codebook (for just the 75 variables in the dataset)](https://docs.google.com/spreadsheets/d/1X7fDKA_BZjSDwUCIo_3fax1viY1_wcJH/edit?usp=sharing&ouid=108879154306217142702&rtpof=true&sd=true)



# Load data and libraries
Load libraries

```{r message = FALSE}

#install.packages("gtsummary", "labelled", "tidyverse")
library(tidyverse)
library(labelled)
library(gtsummary)
```


Load data, make sure either file in same folder as your Rmd or specify correct path. 

```{r}
dat <- readxl::read_xlsx("2022 Fall Data Challenge Dataset.xlsx")
```

Select a few variables to work with

```{r}
dat_small <- dat %>%
  select(SEENJOY, SEGRADES, FHWKHRS, FCSCHOOL)
```

# Summarize data and check for missingness

```{r}
summary(dat_small) 
```

Now I notice some negative levels so I went to the detailed code book and found that means "Valid skip". 

I will recode those to be "NA" before deciding what to do with them. You can choose what to do with missing data - impute or exclude.

```{r}
dat_small[dat_small== -1] <- NA
```

# Make categorical variables into factors

According to code book several of these are categorical variables so we need to recode them so R doesn't interpret them as numeric. Here's a way to do that in bulk, basically either just select all the ones you want categorical or deselect the continuous ones. 

```{r}
# Here I'm mutated across all columns EXCEPT the continuous "FHWKHRS" and making them factor variables

dat_small <- dat_small %>%
  mutate(
    across(
      .cols = !c(FHWKHRS), as.factor
    )
  ) %>%
  drop_na() # for now excluding missing observations

# check on how that did
summary(dat_small)
```



# Labeling variables and factor levels 

For presentation purposes it is often useful to label your variables and factor levels. 


Now lets label the levels using the `forcats` package. 

```{r}

# first make named vectors for the level labels

fcschool_labels <- c(
  "Very satisfied" ="1",
   "Somewhat satisfied" = "2",
   "Somewhat dissatisfied" = "3",
   "Very dissatisfied" = "4"
)

seenjoy_labels <- c(
  "Strongly agree" = "1",
  "Agree" = "2",
  "Disagree" = "3", 
  "Strongly Disagree" = "4"
)

segrades_labels <- c(
  "Mostly A's" = "1", 
  "Mostly B's" = "2",
  "Mostly C's" = "3", 
  "Mostly D's or lower" = "4", 
  "School does not give these grades" = "5"
)

#making new dataframe so we can see difference
dat_new <- dat_small %>%
  mutate(
    FCSCHOOL = fct_recode(FCSCHOOL, !!!fcschool_labels),
    SEENJOY = fct_recode(SEENJOY, !!!seenjoy_labels), 
    SEGRADES = fct_recode(SEGRADES, !!!segrades_labels)
  )

```

And label the column titles using the `labelled` packages. 

```{r}


var_labels <- c(
  SEENJOY = "Enjoyment of school",
  SEGRADES = "Grades",
  FHWKHRS = "Hours spent doing homework",
  FCSCHOOL = "Satisfaction with school"
)


dat_new <- set_variable_labels(dat_new, .labels = var_labels)
```

# Pretty Table 1

Now we can make a pretty table summarizing the data. 

```{r}
dat_new %>%
  tbl_summary(by = SEENJOY) %>%
  add_p() %>%
  modify_spanning_header(all_stat_cols() ~ "**Do you enjoy school?**")
```

# Initial visualizations

```{r}
# set ggplot theme
theme_set(theme_bw())

ggplot(dat_new,
       aes(x = FCSCHOOL)) + 
  geom_bar() + 
  labs(x = "Satisfaction with school")

ggplot(
  dat_new,
  aes(x = SEENJOY,
      y = FHWKHRS)) +
  geom_boxplot()

counts <- table(dat_new$SEENJOY, dat_new$SEGRADES)
mosaicplot(counts, 
           xlab = "Enjoys School", 
           ylab = "Grades",
           main = "Enjoyment by grades")
```




